---
title: 'STAT 428: Homework 2: <br> Random Number Generation and Sampling'
author: "Fu, Xingyu, xingyuf2 <br>Collaborated with: Zhang, Jieyu, jieyuz2"
date: "Due Week 4 Saturday, September 22 by 11.59 PM on Compass"
output:
  html_document:
    theme: readable
    toc: yes
  pdf_document:
    toc: yes
---


```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80)
```
---------------------------------------

Please refer to the [**detailed homework policy document**](https://compass2g.illinois.edu/bbcswebdav/pid-3348982-dt-content-rid-35374883_1/courses/stat_428_120188_166591/1_Info/Syllabus/homework_policy.html) on Course Page for information about homework formatting, submission, and grading. 

---------------------------------------

## Solution 1
**R Syntax and class-typing.**

**(a)** For each of the following commands, either explain why they should be errors, or explain the non-erroneous result.

```
vector1 <- c("3", "3", "8", "2", "4")
vector1*2
sort(vector1)
sum(vector1)
```
1st line: Non-erroneous. The line assigns a vector of characters to variable: vector1.<br>
2nd line: Error. Each element in vector1 is of character type which cannot operate numeric argument, and cannot be multiplied by 2. <br>
3nd line: Non-erroneous. It sorts vector1 in lexicographical order with result as: "2" "3" "3" "4" "8".<br>
4rd line: Error. Each element in vector1 is of character type which cannot operate numeric argument, and cannot be added together. <br>

**(b)** For the next series of commands, either explain their results, or why they should produce errors.

```
vector2 <- c("5",7,12)
vector2[2] + vector2[3]

dataframe3 <- data.frame(z1=5,z2="cat",z3=1)
dataframe3[1,1] + dataframe3[1,3]

list4 <- list(z1="100", z2=25, z3="98", z4=126)
list4[[1]]+list4[[2]]
list4[[2]]+list4[[4]]
list4[2]+list4[4]
list4[2]$z2 + list4[4]$z4

1st line: Non-erroneous. The line assigns a vector of characters to variable: vector2.<br>
2nd line: Error. Each element in vector2 is of character type which cannot operate numeric argument, and cannot be added together. <br>
3nd line: Non-erroneous. It creates a data frame object that has elements of an int, a character, and an int.  <br>
4th line: Non-erroneous. It adds the first and third element (5 and 1) in the data frame object, and thus has result as 6. <br>
5th line: Non-erroneous. It assigns a list that has 4 elements of different types to the variable: list4. <br>
6th line: Error. "100" is character type and 25 is int type. So they cannot be added together.   <br>
7th line: Non-erroneous. This line adds int values 25 and 126 and has result 151. <br>
8th line: Error. `list[2]` and `list[4]` are lists, so they cannot be added together numerically. <br>
9th line: Non-erroneous. It addes up int values in lists `list[2]` and `list[4]`and has result as 151. <br>

## Solution 2
**Working with functions and operators.**
    
**(a)** The colon operator will create a sequence of integers in order. It is a special case of the function `seq()`. Using the help command `?seq` to learn about the function, design an expression that will give you the sequence of numbers from 1 to 10000 in increments of 203. Design another that will give you a sequence between 1 and 10000 that is exactly 18 numbers in length.
```{r, echo=T}
seq(from = 1, to = 10000, by = 203)
seq(from = 1, to = 10000, length.out = 18)
```

**(b)** The function `rep()` repeats a vector some number of times. Explain the difference between `rep(1:4, times=4)` and `rep(1:4, each=4)`.
The first prints 1 to 4 for 4 times It has output 1 2 3 4 1 2 3 4 1 2 3 4 1 2 3 4 <br>
The second prints from 1 to 4, for each number it prints 4 times. It has output 1 1 1 1 2 2 2 2 3 3 3 3 4 4 4 4<br>

## Solution 3
Writing R functions.
Write an `R` function that accepts a vector as an argument and returns the standard deviation of that vector. You may not use the pre-existing functions sd(), var(), or anything similar.
```{r, echo=T}
# a is a input vector, the function will return the standard deviation of it
sd <- function(a) {
  sd = sqrt(sum((a-mean(a))^2/(length(a)-1)))
  return(sd)
}

```

## Solution 4

**Simulating Beta(2, 4) Random Variables: Which method is better?**

In this exercise, you will simulate Beta(2,4) random variables with target density
\[
f(x) \propto x \; (1-x)^{3}.
\]
using the Accept/Reject Algorithm with two different instrumental densities.

- **[Method I]** Use uniform RV's with instrumental density
   \[
   g_{1}(x) = 1,\quad 0 < x < 1.
   \]
   
    - you may use $M \approx 0.1055$.
    - to simulate from $g_{1}$ use the command `x <- runif(1)`.

- **[Method II]** Use the average of two uniform RV's with instrumental density
   \[
   g_{2}(x) = 2 - |4x - 2|,\quad 0 < x <1.
   \]
   
    - you may use $M \approx 0.2499$.
    - the `R` function for absolute value is `abs`.
    - to simulate from $g_{2}$ use the command `x <- mean(runif(2))`.


- **[Question]**  Which method is better? Answer this by responding to the following parts.


- **(a)** Write a program that simulates 10,000 Beta(2,4) random variables using both methods.
- **(b)** Compare the acceptance rates and decide which one is the better algorithm.
- **(c)** Is it correct to use theoretical acceptance rate $\frac{1}{M}$. Why or why not?
- **(c)** Graph histograms of your results.
- **(d)** Summarizes your findings.


- **Hints:**

    - Modify code for simulating Normals from Cauchy that was done in class. 
    - It may help to write two separate programs for each method.

- **(a)** 
```{r, echo=T}

algorithm <- function(n){
  M <- 0.1055
  accept <- 0; num <- 0
  z <- rep(0, times = n)
  while(accept < n){
      x <- runif(1)
      f <- x * (1 - x)^3
      u <- runif(1)
      g <- 1
      if (u <= f/M/g){
        accept <- accept + 1
        z[accept] <- x
      }
      num <- num + 1
    }
list(z = z, accept = n/num)
}
method1 <- algorithm(10000)
method1$accept
```

```{r, echo=T}
algorithm <- function(n){
  M <- 0.2499
  accept <- 0; num <- 0
  z <- rep(0, times = n)
  while(accept < n){
      x <- mean(runif(2))
      f <- x * (1 - x)^3
      u <- runif(1)
      g <- 2 - abs(4 * x - 2)
      if (u <= f/M/g){
        accept <- accept + 1
        z[accept] <- x
      }
      num <- num + 1
    }
list(z = z, accept = n/num)
}
method2 <- algorithm(10000)
method2$accept
```

- **(b)** 
The first one has higher acceptance rate. The first one is better.  

- **(c)** 
No. Since f(x) is only proportional to $x(1 - x)^3$, We could not have the constant and get upper bound as $M'$ , then the $P(acceptance)$ is no longer $\frac{1}{M'}$.  

- **(d)**

```{r, echo=T}
hist(method1$z, freq = FALSE, breaks = 30, main = "First Algorithm Result", xlab = "Accepted Samples")
lines(density(method1$z))

hist(method2$z, freq = FALSE, breaks = 30, main = "Second Algorithm Result", xlab = "Accepted Samples")
lines(density(method2$z))
```

- **(e)** Summarizes your findings.
I think both algorithms simulates the density well, though they may differ in acceptance rate. 

## Solution 5
**Estimation of $\pi$.** $\pi$ is an irrational number between 3 and 4. In `R`, you can get an approximate value of `pi` directly when you enter `pi`. We will now estimate this value ourselves with simulation and compare it with the already available approximation.

- **(a)** Simulate 10000 pairs of random variables $(X_1, X_2)$, both from $Unif(-1,1)$. *Hint: Use the function `runif`. You can generate $X_1$ and $X_2$ separately.*
- **(b)** Find the number of times that $(X_1^2 + X_2^2) \leq 1$ for your sample. Call this number `n`.
- **(c)** Estimate $\pi$ by the number $4n/10000$. Check how close it is to the value given by `pi` in `R`.
- **(d)** Repeat this process 100 times to get a vector of estimates $\tilde{\pi}$.
- **(e)** Estimate the error in our approximation by calculating standard deviation of the vector $\tilde{\pi}$ divided by 10.

- **(a)**
```{r echo=T}
X1=runif(10000,-1,1)
X2=runif(10000,-1,1)
pairs <- data.frame(X1, X2)
```

- **(b)**
```{r echo=T}
n <- dim(subset(pairs, (X1^2+X2^2) <= 1))[1]
n
```

- **(c)**
```{r echo=T}
estimate=4*n/10000
estimate
abs(pi - estimate)
```

- **(d)**
```{r, echo=T}

estimates <- rep(0, 100)
for (i in 1 : 100) {
  X1 <- runif(10000, -1, 1)
  X2 <- runif(10000, -1, 1)
  pairs <- data.frame(X1, X2)
  n <- dim(subset(pairs, (X1^2+X2^2) <= 1))[1]
  estimates[i] <- 4*n/10000
}
estimates
```

- **(e)**
```{r, echo=T}
error = sd(estimates) / 10
error
```

## Solution 6
Suppose we want to generate random samples from a distribution with the following probability density function
$$
f(x) = \begin{cases}
x,    &0\leq x \leq 1,\\
2-x,  &1<x \leq 2.
\end{cases}
$$
Derive the probability inverse transformation $F^{-1}(U)$ and use the inverse transform method to simulate 1000 random samples from this distribution. Graph the density histogram of the sample.

$$ 
F^{-1}(u) = \begin{cases}
\sqrt{2u},          & 0 \leq u \leq \frac{1}{2} \\
2 - \sqrt{2 - 2u},  & \frac{1}{2} < u \leq 1
\end{cases}
$$

``` {r, echo = T}
inverse_F = function(y){
    return(ifelse(y <= 1/2, sqrt(2 * y), 2 - sqrt(2 - 2*y)))
}

us = runif(1000)
samples = sapply(us, inverse_F)
```

``` {r, echo = T}
hist(samples, freq = FALSE, breaks = 30, main = 'Density histogram of the samples', xlab = "x")
y <- seq(0, 2, by = 0.01)
lines(y, sapply(y, function(x) ifelse(x <= 1, x, 2-x)))
```

## Solution 7-10
Do the following exercises from the book: 3.2, 3.3, 3.7, 3.9

## Solution 7 (3.2)
```{r, echo=T}
u <- runif(1000)
f <- log(2 * u) * (u <= 0.5) - log(2 * (1 - u)) * (u > 0.5)

hist(f, probability = T, xlim=c(-5,5), ylim=c(0,0.5), breaks = 40, main="Density histogram of the samples", xlab = "x")
y <- seq(-5, 5, 0.01)
lines(y, 0.5 * exp(-abs(y)))
```

## Solution 8 (3.3)
```{r, echo=T}
u <- runif(1000)
f <- 2/sqrt(1-u)
hist(f, probability = T, breaks=200, xlim=c(2,8), ylim=c(0,1), main="Density histogram of the samples", xlab = "x")
y <- seq(2, 8, 0.01)
lines(y, 8 / (y^3))
```

## Solution 9 (3.7)

```{r, echo=T}
accept <- 0
num <- 0
f <- rep(0, 100)
while (accept < 1000){
  u <- runif(1)
  num <- num + 1
  x <- sqrt(runif(1))
  if (x * (1-x) > u) {
    accept <- accept + 1
    f[accept] <- x
  }
}
hist(f, probability = T, main="Density histogram of the samples", xlab = "x", xlim=c(0,1), ylim=c(0,2), breaks = 30)
y <- seq(0,1,0.01)
lines(y, 12*(y^2)*(1-y))
```


## Solution 10 (3.9)

``` {r, echo = T}
u1 = runif(5000, -1, 1)
u2 = runif(5000, -1, 1)
u3 = runif(5000, -1, 1)
data = ifelse(abs(u3) >= abs(u2) & abs(u3) >= abs(u1), u2, u3)
hist(data, freq = F, main="Density histogram of the samples", xlab = "x", breaks = seq(-1, 1, by = 0.05))
y = seq(-1, 1, by = 0.01)
lines(y, 3/4 * (1 - y^2))
```